---
title: Bookmarking
subtitle: Lecture 30
---

# Deploying

## Several ways to deploy an app

- Option 1: [Posit Connect Cloud](https://connect.posit.cloud/)
- Option 2: [shinyapps.io](https://www.shinyapps.io/) (has a free tier)
- Option 3: [shinylive](https://posit-dev.github.io/r-shinylive/)
- Option 4+: Ad-hoc deployment (e.g. AWS, a server you have, some other hosting service)
  + General rule of thumb: Cheaper = More Effect

# The Basics

## The Problem

-   **State**: Shiny apps are dynamic. Users change inputs to explore data.
-   **Issue**: If you copy the URL, it points to the *initial* state, not the *current* state.
-   **Solution**: **Bookmarking** encodes the current state (inputs) into the URL.

## Enabling Bookmarking

1.  **UI**: Add a `bookmarkButton()`.
2.  **ShinyApp**: Set `enableBookmarking = "url"`.

```{r}
#| eval: false
ui <- function(request) {
  fluidPage(
    textInput("txt", "Text"),
    bookmarkButton()
  )
}

server <- function(input, output, session) { }

shinyApp(ui, server, enableBookmarking = "url")
```

**Note**: `ui` must be a function that takes `request` as an argument.

# Advanced Bookmarking

## Automatic URL Update

-   **Goal**: Update the URL bar *automatically* as inputs change (no button needed).
-   **Server Logic**:
    1.  Observe changes to `input`.
    2.  Call `session$doBookmark()`.
    3.  Use `onBookmarked` to update the query string.

```{r}
#| eval: false
server <- function(input, output, session) {
  # Trigger bookmarking on input change
  observe({
    reactiveValuesToList(input)
    session$doBookmark()
  })
  
  # Update URL in browser
  onBookmarked(updateQueryString)
}
```

## Storing State on Server

-   **`enableBookmarking = "server"`**:
    -   Saves state to a file on the server.
    -   URL becomes short (e.g., `?_state_id=123`).
    -   Useful when state is too large for a URL.

## Storing Custom State

-   Sometimes `input` isn't enough (e.g., you have `reactiveValues` to save).
-   **`onBookmark(function(state) { ... })`**: Save extra data to `state$values`.
-   **`onRestore(function(state) { ... })`**: Read data from `state$values` and restore `reactiveValues`.

# Wrap-Up

## Takeaways

-   **Bookmarking** makes apps shareable.
-   **Function UI**: `ui <- function(request) { ... }` is required.
-   **Modes**:
    -   `"url"`: State in URL (portable, limited size).
    -   `"server"`: State on server (short URL, requires storage).
-   **Automation**: Use `onBookmarked(updateQueryString)` for seamless experience.

## Do Next

1. Read [Chapter 11: Bookmarking](https://mastering-shiny.org/action-bookmark.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. That's it for tonight!
