---
title: Tidy Evaluation
subtitle: Lecture 30
---

# The Problem

## Indirection

-   **Interactive R**: We type `diamonds |> filter(carat > 1)`.
    -   `carat` is a **data-variable** (lives in the dataframe).
-   **Shiny R**: We have `input$var` which contains the string `"carat"`.
    -   `input$var` is an **env-variable** (lives in the environment).
-   **Issue**: `filter(diamonds, input$var > 1)` tries to compare the *string* `"carat"` to 1, which fails or does weird things.

# Data Masking

## The `.data` Pronoun

-   **Solution**: Tell `dplyr` or `ggplot2` exactly where to look.
-   **`.data[[ string ]]`**: Look up the column named `string` in the data frame.
-   **`.env$var`**: Look up the variable `var` in the environment (safer than implicit lookup).

## Example: Filtering

```{r}
#| eval: false
server <- function(input, output, session) {
  
  filtered_data <- reactive({
    # OLD/WRONG: filter(df, input$var > input$min)
    
    # CORRECT:
    df |> filter(.data[[input$var]] > .env$input$min)
  })
  
}
```

## Example: Plotting

-   `aes()` also uses data masking.
-   Use `.data[[ ]]` to map inputs to aesthetics.

```{r}
#| eval: false
output$plot <- renderPlot({
  ggplot(df, aes(.data[[input$x]], .data[[input$y]])) +
    geom_point()
})
```

# Tidy Selection

## Tidy Selectors

-   Functions like `select()`, `pivot_longer()`, `across()` use **tidy selection**.
-   **Interactive**: `select(df, carat, price)`.
-   **Programmatic**: `select(df, all_of(input$cols))`.
    -   `all_of()`: Errors if any column is missing (strict).
    -   `any_of()`: Ignores missing columns (lenient).

```{r}
#| eval: false
# If input$cols is c("carat", "price")
selected_data <- reactive({
  df |> select(all_of(input$cols))
})
```

# Wrap Up

## Rules of Thumb

1.  **Data Masking** (`filter`, `mutate`, `aes`, `arrange`):
    -   Use **`.data[[input$name]]`** to refer to a column by string.
2.  **Tidy Selection** (`select`, `pivot_`, `across`):
    -   Use **`all_of(input$names)`** to refer to columns by a vector of strings.

## Do Next

1. Read [Chapter 12: Tidy Evaluation](https://mastering-shiny.org/action-tidy.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. That's it for tonight!
