---
title: User Feedback
subtitle: Lecture 25
---

# Validation

## Cancelling Execution

-   **Problem**: Missing inputs often cause ugly R errors.
-   **Solution 1**: **`req()`** (Require).
    -   `req(input$id)` stops execution silently if `input$id` is missing, NULL, FALSE, or empty.
    -   Downstream outputs remain blank (or keep old value) instead of crashing.

```{r}
#| eval: false
output$plot <- renderPlot({
  req(input$x) # Stop here if input$x is missing
  plot(input$x)
})
```

## Explicit Validation

-   **Solution 2**: **`validate()`** and **`need()`**.
    -   Checks a condition and displays a specific error message if unsatisfied.

```{r}
#| eval: false
output$table <- renderTable({
  validate(
    need(input$nrows > 0, "Number of rows must be positive")
  )
  head(mtcars, input$nrows)
})
```

# Notifications

## Sending Messages

-   **`showNotification()`**: Displays a temporary message in the corner.
-   **Usage**:
    -   Confirmation ("Saved successfully").
    -   Warnings.
    -   Errors.

```{r}
#| eval: false
observeEvent(input$save, {
  saveData(data())
  showNotification("Data saved successfully!", type = "message")
})
```

# Progress Bars

## Long-Running Tasks

-   Feedback is critical for slow operations.
-   **`withProgress()`**: Wraps a code block.
-   **`incProgress()`**: Increments the bar.

```{r}
#| eval: false
output$plot <- renderPlot({
  withProgress(message = 'Making plot', value = 0, {
    
    incProgress(0.1, detail = "Part 1")
    Sys.sleep(0.5) # Simulate work
    
    incProgress(0.5, detail = "Part 2")
    Sys.sleep(0.5)
    
    plot(rnorm(100))
  })
})
```


## Spinners

- The **`shinycssloaders`** package is a simpler way to add spinners.
- **Setup**: Wrap an output with `withSpinner()` in the UI.
- No server-side changes are needed.

```{r}
#| eval: false
# In UI
library(shinycssloaders)
withSpinner(plotOutput("plot"))

# In server
output$plot <- renderPlot({
  Sys.sleep(2) # Simulate work
  plot(rnorm(100))
})
```



# Confirmation

## Modals

-   Use **`modalDialog()`** for actions that require explicit confirmation (e.g., deleting data).
-   Trigger with `showModal()`.

```{r}
#| eval: false
observeEvent(input$delete, {
  showModal(modalDialog(
    title = "Delete Data?",
    "Are you sure you want to delete this?",
    footer = tagList(
      modalButton("Cancel"),
      actionButton("confirm_delete", "Delete", class = "btn-danger")
    )
  ))
})
```

# Wrap-Up

## Feedback Loops

-   **Silent Rejection**: `req()` (Best for missing inputs).
-   **Explicit Error**: `validate(need())` (Best for wrong inputs).
-   **Status Update**: `showNotification()` (Best for success/info).
-   **Waiting**: `withProgress()` (Best for slow tasks).
-   **Confirmation**: `modalDialog()` (Best for dangerous actions).


## Do Next

1. Read [Chapter 8: User feedback](https://mastering-shiny.org/action-feedback.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. That's it for tonight! See you tomorrow!