---
title: Functions and Iteration
subtitle: Lecture 34
---

# Introduction

## The Motivation

-   **DRY Principle**: Don't Repeat Yourself.
-   **Copy-pasting** is dangerous:
    -   It's easy to make mistakes (copying the wrong variable).
    -   It's hard to update (you have to change code in multiple places).
-   **Solution**:
    1.  **Functions**: Automate common tasks.
    2.  **Iteration**: Apply functions to multiple inputs.

# Functions

## When to write a function?

-   **Rule of Three**:
    -   If you've copied and pasted a block of code more than twice, write a function.
-   **Advantages**:
    -   Given a name that makes it easier to understand.
    -   As requirements change, you only need to update code in one place.
    -   Eliminates the chance of making incidental mistakes when copying and pasting.

## Anatomy of a Function

```{r}
#| eval: false
my_function <- function(arg1, arg2) {
  # Body of the function
  result <- arg1 + arg2
  return(result)
}
```

-   **Name**: Should be descriptive (use verbs).
-   **Arguments**: Inputs to the function (can have defaults).
-   **Body**: The code that executes.
-   **Return**: The output of the function (implicit or explicit).

## Types of Functions {.smaller}

1.  **Vector Functions**:
    -   Input: Vector.
    -   Output: Vector.
    -   Example: `rescale01 <- function(x) { (x - min(x)) / (max(x) - min(x)) }`
2.  **Data Frame Functions**:
    -   Input: Data Frame.
    -   Output: Data Frame (usually).
    -   Example: Wrapping a complex `dplyr` pipeline.
3.  **Plot Functions**:
    -   Input: Data Frame + Parameters.
    -   Output: ggplot object.

# Iteration

## The "R" Way: Vectorization

R is designed for iteration. Many functions work on vectors automatically.

```{r}
x <- 1:5
x * 2  # No loop needed!
```

**Implicit Iteration** should be your first choice.

## Column Iteration: `across()`

Apply a function to multiple columns in a data frame.

```{r}
#| eval: false
library(dplyr)

df %>%
  mutate(
    across(c(x, y, z), ~ .x * 2)
  )
```

-   **Selection**: Select columns by name, type (`where(is.numeric)`), etc.
-   **Function**: Pass the function to apply.

## List Iteration: `purrr::map()`

When you need to iterate over a list or a vector and implicit iteration isn't enough.

-   `map()`: Returns a list.
-   `map_lgl()`, `map_int()`, `map_dbl()`, `map_chr()`: Return an atomic vector.

```{r}
#| eval: false
library(purrr)

path_list <- list("data/file1.csv", "data/file2.csv")
data_list <- map(path_list, read_csv)
```

## Reading Multiple Files

A common data science task:

1.  List files in a directory (`list.files()`).
2.  Read each file (`map()`).
3.  Combine into one data frame (`list_rbind()`).

```{r}
#| eval: false
files <- list.files("data/", pattern = "*.csv", full.names = TRUE)

combined_data <- files %>%
  map(read_csv) %>%
  list_rbind()
```

# Wrap Up

## Summary

-   **Functions** allow you to reuse code logic and avoid copy-paste errors.
-   **Iteration** allows you to apply that logic to many inputs efficiently.
-   **Toolkit**:
    -   `function()` for creating tools.
    -   `dplyr::across()` for data frame columns.
    -   `purrr::map()` for lists and general iteration.

## Do Next

1. Read [Chapter 25 (Functions)](https://r4ds.hadley.nz/functions.html) and [Chapter 26 (Iteration)](https://r4ds.hadley.nz/functions.html) in r4ds..
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. That's it for tonight!