---
title: Workflow
subtitle: Lecture 22
---

# Development Workflow

## Creating and Running

-   **Creation**:
    -   One file app: `app.R`
    -   Two file app: `ui.R` and `server.R`
-   **In RStudio/Positron**:
    -   Click the **Run App** button (play icon) in the toolbar.
-   **Stopping**:
    -   Click the **Stop** sign in the console.

## Shiny Skeleton

```{r}
library(shiny)
ui <- fluidPage(
)
server <- function(input, output, session) {
}
shinyApp(ui, server)
```

# Debugging

## Three Types of Problems

1.  **Unexpected Error**: You get a traceback.
    -   Easiest to fix.
2.  **Incorrect Value**: No error, but wrong result.
    -   Requires investigation (`browser()`).
3.  **Incorrect Logic**: Updates happen at wrong time (or not at all).
    -   Hardest. Unique to Shiny's reactivity.

## Tracebacks and `browser()`

-   **Traceback**: Read from bottom to top to find *your* code.
-   **`browser()`**:
    -   Place `browser()` inside your server function.
    -   Execution pauses when it hits that line.
    -   You can inspect `input`, `output`, and intermediate variables in the console.

```{r}
server <- function(input, output, session) {
  observeEvent(input$go, {
    browser() # Execution pauses here
    # Inspect input$x, calculated values, etc.
  })
}
```

## Print Debugging

-   Classic technique: Print values to see what's happening.
-   **Use `message()`** instead of `print()`.
    -   Goes to standard error (logs).
-   **Use `glue::glue()`** for easy formatting.

```{r}
library(glue)
server <- function(input, output, session) {
  val <- reactive({
    message(glue("Updating val. input$x is {input$x}"))
    input$x * 2
  })
}
```

## Debugging Reactivity

-   Reactivity is **lazy** and **non-linear**.
-   Code doesn't run top-to-bottom.
-   Visualizing the **Reactive Graph** helps.
-   Use `message()` inside `reactive()` to verify:
    1.  Is it running?
    2.  Is it running *too often*?
    3.  Is it running in the *expected order*?

# Getting Help

## The Reprex

-   **Reprex**: **Repr**oducible **Ex**ample.
-   Critical for getting help (StackOverflow, RStudio Community).
-   **Properties**:
    -   **Minimal**: Remove everything not related to the bug.
    -   **Self-contained**: Copy/pasteable. runs on anyone's machine.
        -   No `read.csv("C:/MyFiles/data.csv")`.
        -   Use code to generate dummy data.

## Creating a Shiny Reprex

1.  Isolate the problem (remove unrelated UI/Server code).
2.  Use built-in datasets (e.g., `mtcars`).
3.  Combine UI and Server in one file.

```{r}
library(shiny)

ui <- fluidPage(
  numericInput("n", "N", 5),
  textOutput("txt")
)

server <- function(input, output, session) {
  output$txt <- renderText(input$n * 2)
}

shinyApp(ui, server)
```

## Practice!

Let's work through the case study in the text!

# Summary

## Workflow Tips

-   **Read tracebacks** carefully.
-   Use **`browser()`** to stop and inspect.
-   Use **`message()`** to track reactive execution.
-   When stuck, create a **Reprex**.

## Do Next

1. Read [Chapter 5: Workflow](https://mastering-shiny.org/action-workflow.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. Move on to Lecture 23!