---
title: "ggplot2 layers"
subtitle: "Lecture 7"
execute: 
  message: false
  warning: false
---

```{r}
#| label: setup
#| message: false
#| echo: false

# load packages
library(tidyverse)
library(nycflights13)
library(countdown)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)

# flights <- flights |> 
#   slice_sample(n = 1000)
```

# The layered grammar of graphics

## Data: Flights from NYC in 2013

We will be using the `nycflights13::flights` dataset. This dataset contains information about all flights that departed from New York City (EWR, LGA, JFK) in 2013.

```{r}
glimpse(flights)
```

## A simple scatterplot {.smaller}

 Let's create a simple scatterplot of departure delay vs. arrival delay.

## Adding layers

A plot is composed of layers. Each layer has:

- A **geom**: The geometric object that represents the data (e.g., points, lines, bars).
- An **aesthetic mapping**: How variables are mapped to visual properties.
- A **stat**: A statistical transformation (e.g., counting, summarizing).
- A **position adjustment**: How to arrange elements that would otherwise overlap.

## Adding a layer: Trend line with `geom_smooth()` {.smaller}

Let's add a trend line... What does se = FALSE do?

# Aesthetic Mappings

## Global vs. Local Mappings

-  Aesthetic mappings can be: 
  - Global: specify in `ggplot()` so they apply to all layers
  - Local: specify in each `geom_*()` so they apply only to that layer

## Using `color` as a local aesthetic {.smaller}

Let see what happens when we use `color` as a global aesthetic mapping in `geom_point()`.

How can we make a single color for all points but still have the smooth line?

## Why isn't this plot blue?

```{r}
# This is probably not what you want
ggplot(flights, aes(x = dep_delay, y = arr_delay)) +
  geom_point(aes(color = "blue"))
```

## Setting vs. Mapping

- To apply a constant property to all points, set it outside of `aes()`

```{r}
ggplot(flights, aes(x = dep_delay, y = arr_delay)) +
  geom_point(color = "blue")
```

# Geoms

## Common geoms

- `geom_point()`: Scatterplots
- `geom_line()`: Line charts
- `geom_bar()`: Bar charts
- `geom_histogram()`: Histograms
- `geom_boxplot()`: Boxplots
- `geom_smooth()`: Trend lines
- `geom_density()`: Density plots
- [Many more!](ttps://ggplot2.tidyverse.org/reference/index.html#geom)


# Statistical Transformations

## `geom_bar` vs `geom_col`

- `geom_bar()` uses `stat_count()` by default to count the number of observations for each category

```{r}
#| label: bar-chart

ggplot(data = flights, mapping = aes(x = carrier)) +
  geom_bar()
```

## Changing the `stat` argument

- You can change the statistical transformation used by a geom with the `stat` argument...

```{r}
flights |>
  count(carrier) |> 
  ggplot(mapping = aes(x = carrier, y = n)) +
    geom_bar(stat = "identity")
```

## Changing the stat

- Using `after_stat()` to modify computed variables from a stat

```{r}
ggplot(data = flights) +
  geom_bar(mapping = aes(x = carrier, y = after_stat(prop), group = 1)) +
  labs(y = "Proportion")
```

## Practice

Let's try to overlay a density plot on top of a histogram of `dep_delay`...


# Position Adjustments

## Overplotting

- Overplotting: When many points overlap in a scatterplot, making it hard to see the distribution of points
- Common solution: Use transparency with `alpha` aesthetic
- Another solution: Use position adjustments to spread out points

## Jittering points

```{r}
#| warning: false
#| message: false

library(patchwork)
p1 <- ggplot(flights, aes(x = month, y = dep_delay)) +
  geom_point()
p2 <- ggplot(flights, aes(x = month, y = dep_delay)) +
  geom_jitter()
p1 + p2
```


## Bar chart adjustments

For bar charts, you can use `position` to change how bars are arranged.

- `stack` (default): Stack bars on top of each other.
- `fill`: Stack bars, but standardize to have the same height. Good for comparing proportions.
- `dodge`: Place bars next to each other.

## Bar chart adjustments

```{r}
p1 <- ggplot(flights, aes(x = carrier, fill = origin)) +
  geom_bar(position = "stack")
p2 <- ggplot(flights, aes(x = carrier, fill = origin)) +
  geom_bar(position = "fill")
p3 <- ggplot(flights, aes(x = carrier, fill = origin)) +
  geom_bar(position = "dodge") +
  labs(title = "dodge")
```

## Bar chart: Stacked
```{r}
p1
```

## Bar chart: Standardized
```{r}
p2
```

## Bar chart: Dodged
```{r}
p3
```


# Facets

## `facet_wrap` and `facet_grid` {.smaller}

- Facets allow you to break your plot into subplots based on a categorical variable.
- `facet_wrap`: Creates a series of plots wrapped into a rectangle
- `facet_grid`: Creates a grid of plots based on two categorical variables

## Practice

-  Let's visualize the relationship between `origin`, `carrier`, `dep_delay`, and `arr_delay` using facets!

# Coordinate Systems

## Flipping coordinates

```{r}
p1 <- ggplot(data = flights, mapping = aes(x = carrier)) +
  geom_bar()

p2 <- ggplot(data = flights, mapping = aes(x = carrier)) +
  geom_bar() +
  coord_flip()
p1 + p2
```


## The layered grammar in summary

```r
#| eval: false
`ggplot(data = <DATA>) +`
`  <GEOM_FUNCTION>(`
`    mapping = aes(<AESTHETIC_MAPPINGS>),`
`    stat = <STAT>,`
`    position = <POSITION>`
`  ) +`
`  <COORDINATE_SYSTEM> +`
`  <FACET_FUNCTION>`
```
## Do Next

1. Read [Chapter 9: Layers](https://r4ds.hadley.nz/layers.html) from r4ds.
2. Open the Recitation Gem and say "Provide me practice problems for Chapter 9" or work through some of the exercises in the text.
3. Move on to Lecture 08.
