---
title: Feature Engineering with Recipes
subtitle: Lecture 31
---

# Introduction

## Preprocessing Data

- **Feature Engineering**: transforming variables to make them more suitable for a model (or visualization!).
- Common tasks:
  -   Converting categorical data to numeric (Dummy variables).
  -   Centering and Scaling (Normalization).
  -   Handling missing data.
  -   Transforming skewed distributions (Log).
- **The `recipes` package**: A tidy interface for data preprocessing.

## The `recipe` Concept

- A recipe is a blueprint for data processing.
- It defines **what** you want to do, not **when** to do it.
- **Workflow**:
  1.  `recipe()`: Define the formula and data.
  2.  `step_*()`: Add processing steps.
  3.  `prep()`: Train the recipe (calculate means, SDs, levels, etc.).
  4.  `bake()`: Apply the recipe to new data.

## Load Packages

```{r}
library(tidymodels)
library(palmerpenguins)
library(tidyverse)
```

# Defining a Recipe

## The Basics

- Start with `recipe(formula, data)`.
- **Role**: `outcome` (LHS) vs `predictor` (RHS).
- **Data**: Used only to check variable names and types (not for training yet).

```{r}
simple_rec <- recipe(species ~ ., data = penguins)
simple_rec
```

## Adding Steps

- Steps are added using pipes (`|>`).
- Examples (order matters!):
  -   Impute missing data first.
  -   Individual transformations (log).
  -   Discretization / Dummy Variables.
  -   Normalization (Center/Scale).
  -   Multivariate transformations (PCA).

# The Lifecycle: Prep & Bake

## `prep()`: Estimating Parameters

- `prep()` executes the recipe on the training data.
- It calculates necessary statistics (means, SDs, factor levels).

```{r}
# Define the full recipe
final_rec <- recipe(species ~ ., data = penguins) |>
  step_naomit(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

# Train it
prepped_rec <- prep(final_rec, training = penguins)
prepped_rec
```

## `bake()`: Applying to Data {.smaller}

- `bake()` applies the transformations to data.
- Use `new_data = NULL` to get the processed data you used to `prep` the recipe.

```{r}
# Process the data
processed_data <- bake(prepped_rec, new_data = NULL)

processed_data |> head()
```

## Why `prep` and `bake`?

- **Separation of concerns**: You define the steps once, then apply them to:
  -   Training data (data you use to create your model)
  -   Testing data (data you use to evaluate your model)
  -   New data in production
- Ensures **Data Leakage** prevention:
  -   Means/SDs are calculated on *training* data only.
  -   Applied to test data.

# Common Steps

## Imputation & Transformations

- `step_naomit()`: Remove rows with NA (simple).
- `step_impute_*()`: Impute missing values (mean, median, knn).
- `step_log()`: Log transform skewed variables.
- `step_mutate()`: General mutations (similar to `dplyr::mutate`).

```{r}
# Example: Log transform body mass
rec_log <- recipe(species ~ body_mass_g, data = penguins) |>
  step_log(body_mass_g)
```

## Numeric Scaling

- **Centering**: Subtract the mean (mean = 0).
- **Scaling**: Divide by standard deviation (sd = 1).
- Many important for algorithms sensitive to scale.
- Use `step_normalize()` (does both center and scale).

```{r}
rec_norm <- recipe(species ~ bill_length_mm, data = penguins) |>
  step_normalize(bill_length_mm)
```

## Categorical Data (Dummy Variables)

- Many algorithms requires numeric input (PCA, K-Means, Linear Regression).
- **Dummy Variables**: Convert categories into binary 0/1 columns.
- `step_dummy(all_nominal_predictors())`.

```{r}
rec_dummy <- recipe(body_mass_g ~ species + island, data = penguins) |>
  step_dummy(all_nominal_predictors())
```

- Watch out for the "dummy trap" (multicollinearity), usually handled by default (one level is reference).

## Getting ready for PCA and Clustering

- For **PCA** and **Clustering**:
  -   Use `step_normalize()` (Variable scale influences distance).
  -   Use `step_dummy()` (Must be numeric).
  -   Often drop unwanted variables.

```{r}
clustering_rec <- recipe(~ ., data = penguins) |>
  step_rm(species, year, sex) |> # Remove non-predictive/outcome vars
  step_naomit(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

clustering_rec
```

# Wrap-Up

## Recap

- **Recipe**: A plan for data processing.
- **Steps**:
  -   `step_dummy()`: Categorical -> Numeric.
  -   `step_normalize()`: Center and Scale.
- **Process**:
  1.  Define `recipe()`.
  2.  Add `step_*()`.
  3.  `prep()` on training data.
  4.  `bake()` to get the result.
- **Next Time**: Using this for Dimensionality Reduction (PCA)!


## Do Next

1. Read [Chapter 8: Feature engineering with `recipes`](https://www.tmwr.org/recipes) from Tidy Modeling with R.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. Move on to Lecture 32!
