---
title: "Regular Expressions"
subtitle: "Lecture 13"
---

```{r}
#| label: setup
#| message: false
#| echo: false

# load packages
library(tidyverse)
library(countdown)
library(stringr)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)
```

# Introduction

## Regular Expressions (Regex)

- A powerful tool for describing patterns in strings.
- Used for finding, extracting, and replacing text.
- `stringr` provides a consistent interface for working with regex.

# Key Functions

## `str_detect()`

- `str_detect()` returns `TRUE` if a pattern is found in a string, `FALSE` otherwise.

```{r}
#| echo: true
x <- c("apple", "banana", "pear", "pineapple", "naartjie")
str_detect(x, "an")
```

## `str_count()`

- `str_count()` counts the number of matches in a string.

```{r}
#| echo: true
str_count("banana", "a")
```

## `str_extract()` and `str_extract_all()`

- `str_extract()` extracts the first match.
- `str_extract_all()` extracts all matches.

```{r}
#| echo: true
str_extract("banana", "an")
str_extract_all("banana", "an")
```

## `str_replace()` and `str_replace_all()`

- Replaces the first or all matches with a new string.

```{r}
#| echo: true
x <- c("Dr_Eric_Friedlander", "Dr_Brandy_Wiegers", "Anthony_Campitelli")
str_replace(x, "_", " ")
str_replace_all(x, " ", " ")
```

## Regex

Instead of inputting literal strings, you can use regex patterns to describe more complex matches.

```{r}
#| echo: true
x <- c("Dr_Eric_Friedlander", "Dr_Brandy_Wiegers", "Anthony_Campitelli")
# remove all vowels
str_replace_all(x, "[aeiouAEIOU]", "")
```

# Pattern Components

## Anchors

- `^` matches the start of the string.
- `$` matches the end of the string.

```{r}
#| echo: true
x <- c("apple", "banana", "pear")
str_detect(x, "^a")
str_detect(x, "a$")
```

## Character Classes

- `.` matches any character except a newline.
- `\d` matches any digit.
- `\s` matches any whitespace.
- `[abc]` matches a, b, or c.
- `[^abc]` matches anything except a, b, or c.
- `|` matches either the expression before or after the `|`.

## Repetition

- `?`: 0 or 1 time.
- `+`: 1 or more times.
- `*`: 0 or more times.
- `{n}`: exactly n times.
- `{n,}`: n or more times.
- `{n,m}`: between n and m times.

## Practice!

What do each of these do?

- `"^b.*a$"`
- `"^.{5}$"`
- `"[aeiou]"`
- Create a regular expression that will match telephone numbers as commonly written in your country.

# Grouping and Back References

## `()` for Grouping

- Parentheses create a "capturing group" to extract parts of a match.

```{r}
#| echo: true
str_match("apple, banana, pear", "([a-z]+), ([a-z]+)")
```

## Back References

- `\1`, `\2`, etc. refer to previously captured groups.

```{r}
#| echo: true
str_replace("abab", "(a)(b)", "\\2\\1")
```

## Practice!

Exercise 6 from 15.4.7.

# Other Tools

## `tidyr::separate_wider_regex()`

- Separates a column into multiple columns using regex with capture groups.

```{r}
#| echo: true
df <- tibble(x = "123-abc")
df |> separate_wider_regex(x, c(num = "\\d+", "-", chr = "[a-z]+"))
```

## `fixed()`

- Use `fixed()` to match a literal string without interpreting it as a regex.

```{r}
#| echo: true
str_detect("a.b", fixed("."))
```

## Practice!

Let's extract the AR codes and the county names from the Naturalization data!

# Wrap-Up


## Do Next

1. Read [Chapter 15: Regular Expression](https://r4ds.hadley.nz/regexps.html) from r4ds.
2. Open the Recitation Gem and say "Provide me practice problems for Chapter 15" or work through some of the exercises in the text.
3. That it for tonight! See you tomorrow.
