---
title: "Page Layout"
subtitle: "Mastering Shiny, Chapter 6"
---

# Single Page layout

## `fluidPage()`

- The foundation of most Shiny app layouts.
- Creates a "fluid" page that automatically adjusts to the browser window size.
- All UI elements are typically placed inside `fluidPage()`.

```{r}
#| eval: false
ui <- fluidPage(
  "Hello, world!"
)
```

## `titlePanel()`

- A common first element in `fluidPage()` is `titlePanel()`.
- It creates a large heading, suitable for the app's title.
- It also sets the window title of the browser.

```{r}
#| eval: false
ui <- fluidPage(
  titlePanel("My Shiny App")
)
```

## `sidebarLayout()`

- A classic two-column layout.
- Consists of `sidebarPanel()` (for inputs) and `mainPanel()` (for outputs).
- `sidebarPanel()` is typically placed on the left.

```{r}
#| eval: false
ui <- fluidPage(
  titlePanel("My App"),
  sidebarLayout(
    sidebarPanel(
      # inputs
    ),
    mainPanel(
      # outputs
    )
  )
)
```

## `sidebarLayout()` positions

- You can control the position of the sidebar with the `position` argument.
- `position = "left"` (default)
- `position = "right"`

```{r}
#| eval: false
sidebarLayout(
  sidebarPanel(
    # inputs
  ),
  mainPanel(
    # outputs
  ),
  position = "right"
)
```

## Multi-row Layouts with `fluidRow()` {.smaller}

- For more complex layouts, you can build your own grid.
- `fluidRow()` creates a row.
- `column()` creates columns within a row.
- The total width of columns in a row must be 12.

```{r}
#| eval: false
fluidPage(
  fluidRow(
    column(4, 
      "First column"
    ),
    column(8, 
      "Second column"
    )
  ),
  fluidRow(
    column(6, 
      "Third column"
    ),
    column(6, 
      "Fourth column"
    )
  )
)
```

## Nesting Layouts

- You can nest `fluidRow()` and `column()` to create more complex layouts.
- You can also nest `sidebarLayout()` inside a `column()`.

# `tabsetPanel()` and `navlistPanel()`

## `tabsetPanel()`

- Useful for breaking down a complex app into smaller pieces.
- Creates a set of tabs.
- Each tab is created with `tabPanel()`.

```{r}
#| eval: false
mainPanel(
  tabsetPanel(
    tabPanel("Plot", plotOutput("plot")), 
    tabPanel("Summary", verbatimTextOutput("summary")), 
    tabPanel("Table", tableOutput("table"))
  )
)
```

## `tabsetPanel()` options

- `id`: Assign an ID to get the currently selected tab in the `server`.
- `selected`: Set the initially selected tab.
- `type`: Change the appearance of the tabs (e.g., "tabs", "pills").
- `position`: Place tabs on "top", "bottom", "left", or "right".

## `navlistPanel()`

- An alternative to `tabsetPanel()`.
- Renders tabs as a sidebar navigation list.
- Well-suited for a sequence of linear steps.
- Cannot be used inside `sidebarLayout`.

```{r}
#| eval: false
navlistPanel(
  "Header",
  tabPanel("First", "First content"),
  tabPanel("Second", "Second content"),
  tabPanel("Third", "Third content")
)
```

# `navbarPage()`

## Page-level navigation

- For multi-page apps, `navbarPage()` is the tool of choice.
- It creates a navigation bar at the top of the page.
- It replaces `fluidPage()` as the top-level UI function.
- Each `tabPanel` becomes a "page".

```{r}
#| eval: false
ui <- navbarPage(
  "My App",
  tabPanel("Page 1", "Content for page 1"),
  tabPanel("Page 2", "Content for page 2")
)
```

## Menus in `navbarPage()`

- You can group related pages into a menu using `navbarMenu()`.
- This is useful for organizing a large number of pages.

```{r}
#| eval: false
navbarPage(
  "My App",
  tabPanel("Page 1", ),
  tabPanel("Page 2", ),
  navbarMenu("More",
    tabPanel("Sub-page A", ),
    tabPanel("Sub-page B", )
  )
)
```

## `navbarPage()` options

- `title`: The title of the app, shown in the navbar.
- `id`: To access the selected tab from the server.
- `collapsible = TRUE`: The navbar collapses on small screens.
- `theme`: To apply a custom theme.

# Themes

## Getting started with `bslib`

- Shiny's appearance is controlled by the Bootstrap CSS framework.
- The `bslib` package makes it easy to customize Bootstrap themes.
- Use `bs_theme()` to create a theme and pass it to the `theme` argument of the page layout function.

```{r}
#| eval: false
ui <- fluidPage(
  theme = bslib::bs_theme(bootswatch = "darkly"),
)
```

## Real-time theming

- `bslib` offers an interactive theming tool.
- Run `bs_theme_preview()` on your UI to launch a special version of your app where you can tweak theme settings in real-time.

```{r}
#| eval: false
# In the R console
my_ui <- fluidPage()
bslib::bs_theme_preview(my_ui)
```

## Customizing components

- `bs_theme()` allows you to customize many aspects of the theme.
  - `bg`, `fg`: Background and foreground colors.
  - `primary`, `secondary`: Accent colors.
  - `base_font`, `heading_font`, `code_font`: Fonts.

```{r}
#| eval: false
my_theme <- bslib::bs_theme(
  bg = "#202123",
  fg = "#B8BCC2",
  primary = "#EA80FC",
  base_font = bslib::font_google("Inter"),
  "font-size-base" = "1.1rem"
)
```

# The Underlying HTML

## HTML Basics

- Shiny UIs are converted into HTML, CSS, and JavaScript.
- HTML (HyperText Markup Language) provides the structure.
- Tags (`<tag>`) create elements. They can have attributes (`<tag attribute="value">`).
- Tags are nested to form a hierarchy.

```html
<p>A <strong>very</strong> important paragraph.</p>
```

## Using HTML in Shiny

- Shiny functions like `p()`, `h1()`, `strong()` create common HTML tags.
- For any tag, you can use `tags$<tag>()`.
- All tag functions take child elements and named attributes.

```{r}
#| eval: false
tags$p(
  "A ",
  tags$strong("very"),
  " important paragraph."
)
```

## Including CSS

- CSS (Cascading Style Sheets) controls the appearance.
- Use the `includeCSS()` function to include a custom CSS file.
- Place your CSS file in a `www` subdirectory of your app.

```{r}
#| eval: false
# In ui.R
fluidPage(
  includeCSS("www/styles.css"),
)

/* In www/styles.css */
body {
  color: purple;
}
```

## When to use raw HTML

- When you need a tag that Shiny doesn't provide a function for.
- For minor tweaks to the UI.
- To integrate with other JavaScript libraries that require a specific HTML structure.
- But for major style changes, prefer `bslib`.

# Wrap Up

## Do Next

1. Read [Chapter 6: Layouts, Themes, and HTML](https://mastering-shiny.org/action-layout.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. Move on to Lecture 24!