---
title: "Factors"
subtitle: "Lecture 14"
---

```{r}
#| label: setup
#| message: false
#| echo: false

# load packages
library(tidyverse)
library(countdown)
library(forcats)
library(patchwork)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)
```

# Introduction

## Categorical Data and Factors

- Factors are R's data structure for categorical variables.
- The `forcats` package (part of `tidyverse`) provides a suite of tools for working with factors.
- Useful for reordering and relabeling for better visualizations.

# Creating Factors

## `factor()` and `fct()`

- Create factors from a character vector.
- `levels()` shows the categories of the factor.
- The levels have an order (not to be confused with "ordered factors" which are different)

```{r}
#| echo: true
x <- c("Male", "Female", "Female", "Male")
f <- factor(x)
f
levels(f)
```

## Data: Idaho Naturalization

- We're going to use Naturalization data from the Idaho State Archives again
- This dataset was assembled by Dr. Rachel Miller
- [Link to Google Sheet](https://docs.google.com/spreadsheets/d/1UpMPcO8Jr3FLlFvy27a5xNMcCPqF20wur8YtOjH8tw0/edit?gid=1474309575#gid=1474309575)
- Let's load the data and convert the country of origin to a factor

# Modifying Factor Order

## `fct_reorder()`

- Reorders a factor by another variable, useful for plots.
- For example, ordering bars in a bar chart.
- Let's create a bar chart of country of origin ordered by the average year of arrival.

## `fct_infreq()`

- Reorders a factor by the frequency of its values.
- Let's create a bar chart of country of origina by the frequency

## `fct_relevel()`

- Manually move levels to the front.
- Let's move "Spain" to the front.

## Modifying Factor Levels

- `fct_recode()`: Changes the names of factor levels.
- `fct_collapse()`: Combines several levels into one.
- `fct_lump_*()`: lump together the least/most frequent levels into an "Other" category.

## Practice

- Let's make the country of origin factor more manageable by lumping together the least frequent countries into "Other" and recoding some country names for clarity.

## Ordered Factors

- A factor that enforces a strict ordering of levels
- Useful for ordinal data (e.g., ratings)
- This class: only really matters if you tie it to a color aesthetic
  + Will select a color scheme that implies an ordering

# Wrap Up

## Do Next

1. Read [Chapter 16: Factors](https://r4ds.hadley.nz/factors.html) from r4ds.
2. Open the Recitation Gem and say "Provide me practice problems for Chapter 16" or work through some of the exercises in the text.
3. Move on to lecture 15.

