---
title: Dynamic UI
subtitle: Lecture 28
---

# Updating Inputs

## The `update*` Family

-   **Goal**: Change an input's state (value, choices, label) from the server.
-   **Functions**: `updateTextInput()`, `updateSliderInput()`, `updateSelectInput()`, etc.
-   **Mechanism**:
    -   User action triggers an `observeEvent()`.
    -   Server sends a message to the browser to update the specific control.

```{r}
#| eval: false
observeEvent(input$reset, {
  updateSliderInput(inputId = "s1", value = 0)
})
```

## Hierarchical Selects {.smaller}

Consider this scenario from Chapter 9 of the textbook. 

> Create an app that lets the user upload a csv file, select one variable, draw a histogram, and then download the histogram. For an additional challenge, allow the user to select from .png, .pdf, and .svg output formats.

# Dynamic Visibility

## Hiding and Showing {.smaller}

-   **Goal**: Show different UI elements based on user actions.
-   **Technique**: Hidden Tabsets.
    -   `tabsetPanel(id = "wizard", type = "hidden", ...)`
    -   `updateTabsetPanel(inputId = "wizard", selected = "panel2")`

```{r}
#| eval: false
ui <- fluidPage(
  actionButton("go_next", "Next"),
  tabsetPanel(id = "wizard", type = "hidden",
    tabPanel("page1", "First step content..."),
    tabPanel("page2", "Second step content...")
  )
)

server <- function(input, output, session) {
  observeEvent(input$go_next, {
    updateTabsetPanel(inputId = "wizard", selected = "page2")
  })
}
```

# Generative UI

## `uiOutput` and `renderUI`

-   **Goal**: Generate HTML or Inputs entirely from Python/R code.
-   **Usage**:
    1.  **UI**: Place `uiOutput("id")` where you want the content.
    2.  **Server**: Assign `renderUI({ ... })` to `output$id`.
    3.  **Return**: The `renderUI` block must return Shiny tags (e.g., `tagList`, `div`, inputs).

## Circular References

-   **Warning**: Be careful when creating inputs that trigger the same code that created them.
-   **Loop**: `renderUI` -> creates input -> input changes -> triggers `renderUI` -> re-creates input.
-   Rate limiting or logic checks are often needed.

# Wrap Up

## Levels of Dynamism

1.  **Update**: Modify existing inputs (`update*Input`).
2.  **Visibility**: Show/Hide pre-defined sections (`tabsetPanel`).
3.  **Generation**: Create new UI from scratch (`uiOutput`/`renderUI`).

## Do Next

1. Read [Chapter 10: Dynamic UI](https://mastering-shiny.org/action-dynamic.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. Move on to Lecture 29!