---
title: Our First Analysis
format: revealjs
auto-stretch: false
---

## Today

-   Today we're going to split up into groups and each analyze a data set of our choice
-   We will take turns presenting on what we've done
-   After each presentation, we'll have a brief discussion where we critique each other's analysis

## \#TidyTuesday

-   TidyTuesday is a weekly data project aimed at the R community
-   Each week a new data set is released along with a short description
-   People analyze the data set and share their work on social media using the hashtag #TidyTuesday
-   You can find the data sets and more information at the [TidyTuesday GitHub repository](https://github.com/rfordatascience/tidytuesday)

## Today {.smaller}

-   Each group will pick a data set from the TidyTuesday repository
-   Each group will *do their best* to analyze that data
-   Advice:
    -   Try to choose a more recent data set, and one that already has questions associated with it
    -   There is usually an import script at the bottom of each dataset
    -   It is likely that you will need to install some packages to work with the data
    -   Don't worry about being perfect
    -   Focus on exploring the data and telling a story with it
    -   Ask for help when you need it

# Groups

## Group Roles

-   Coder: responsible for typing for your group
-   Reporter: responsible for sharing your group's work with the class
-   Researcher: responsible for looking up any questions your group has
-   Everyone should contribute to the discussion and help with writing the code
-   There will be several parts of this activity, between each part, rotate roles
    -   Coder -> Reporter -> Researcher -> Coder ...
    -   For now, old Coder copy and paste script in Teams chat to new Coder

## Get in your groups {.smaller}

-   You can find a spreadsheet in the Jan 6th Activities Module in the Classword section on Teams with your team assignments
-   Get in your groups now
-   Introduce yourselves
-   Decide who will be the first Coder, Reporter, and Researcher
-   Decide on a group name
-   Create a post in the Activities Channel on Teams with all of this information

```{r}
countdown::countdown(minutes = 5)
```

## Selecting a Dataset {.smaller}

-   Do the following:
    +   Find the data set on the TidyTuesday GitHub repository
    +   Load that data set into R
    +   Post a link to the dataset as a response to your group's post in the Activities Channel on Teams
-   Answer the following questions (be prepared to give a 2-minute report to the class):
    +   What is the data about?
    +   What question(s) are you interested in answering with this data?
    +   How many data frames are there and how many rows and columns does it have?
    +   What data frames and variables are you going to focus on?

```{r}
countdown::countdown(minutes = 20)
```

## Explore the Data {.smaller}

-  Be prepared to present the following (less than 5 minutes per group):
    +   Create up to 5 visualizations or tables that help answer your questions
    +   Explain what each visualization or table shows
    +   Explain what you learned from each visualization or table
    +   Explain any challenges you had while working with the data
    +   Be prepared to answer questions and explain your code if you are asked
-   Copy and paste your entire script (up to this point) into Teams as a response to your group's post

```{r}
countdown::countdown(minutes = 20)
```

## Time Permitting: Refine and Extend {.smaller}

-   If you have time left, refine your visualizations and analyses based on our discussions

## Wrap Up {.smaller}

-   Before you leave, complete your reflection on Teams!

