---
title: More Analysis
format: revealjs
auto-stretch: false
---

## Today

-   Return to TidyTuesday analysis from last class
-   Work with some real experimental data generated from by Dr. Belissimo's class

## Group Roles: Reminder {.smaller}

-   Coder: responsible for typing for your group
-   Reporter: responsible for sharing your group's work with the class
-   Researcher: responsible for looking up any questions your group has
-   Everyone should contribute to the discussion and help with writing the code
-   There will be several parts of this activity, between each part, rotate roles
    -   Coder -> Reporter -> Researcher -> Coder ...
    -   For now, old Coder copy and paste script in Teams chat to new Coder

## Groups for Today

-   If you didn't present yesterday, same groups as yesterday
-   If you did present yesterday, you've been assigned a new group (check Teams)


# Back to TidyTuesday

-   Pokemon: Refine your visualizations using pivoting (be prepared to present to class)
-   DnD: Refine your visualizations using pivoting (be prepared to present to class)
-   Everyone else: Work on the challenges below (you may use any order you like)
    +   Do not use AI to generate, you may use AI to debug code
    +   Upload completed image once complete by pasting into replay in Activities Channel

```{r}
#| echo: false

countdown::countdown(minutes = 30)
```

## Challenge 1A: Parks

- Use the `parks.csv` found in Teams to make this plot.

```{r}
#| echo: false
library(tidyverse)
parks <- read_csv(here::here("Data", "Parks", "parks.csv"))

parks |> 
    filter(city %in% c("New York", "Chicago")) |> 
    ggplot(aes(x = year, y = total_points, fill = city)) +
    geom_col(position="dodge")
```

## Challenge 1B: Parks

- Use the `parks.csv` found in Teams to make this plot.

```{r}
#| echo: false

parks |> 
    filter(city %in% c("New York", "Chicago")) |> 
    select(year, city, total_points) |>
    pivot_wider(
        names_from = city,
        values_from = total_points
    ) |> 
    ggplot(aes(x = `New York`, y = Chicago)) +
    geom_point() + 
    labs(
        x = "New York Total Points",
        y = "Chicago Total Points"
    )
```

## Challenge 2: Valentine's Day

[Generate this plot (without usiung AI):](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-02-13/readme.md)

```{r}
#| echo: false

library(tidyverse)
historical_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/historical_spending.csv')

historical_spending |> 
    select(-PercentCelebrating) |> 
    pivot_longer(
        cols = !Year,
        names_to = "Category",
        values_to = "Spending"
    ) |> 
    ggplot(aes(x = Year, y = Spending, color = Category)) +
        geom_line()
```

## Challenge 3: Pokemon

[Generate the following data frame from the pokemon data (no AI):](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-04-01/readme.md)
```{r}
#| echo: false

library(knitr)
pokemon_df <- pokemon::pokemon

pokemon_df |> 
    pivot_longer(
        cols = type_1:type_2,
        names_to = "Type_num",
        values_to = "Type"
    ) |> 
    pivot_longer(
        cols = hp:speed,
        names_to = "Stat",
        values_to = "Value"
    ) |>
    group_by(Type, Stat) |> 
    summarize(Avg_Value = mean(Value, na.rm = TRUE), .groups = "drop") |>
    select(Type, Stat, Avg_Value) |> 
    head(10) |> 
    kable()
```

# HHP Data

## Reasearch Question {.smaller}

- Traditional Resistance Training (TRT): Involves performing a set number of repetitions continuously until the set is complete, followed by a rest period.
    +   Example: 3 sets of 10 repetitions with 60 seconds rest between sets.
- Cluster Resistance Training (CRT): Involves breaking a set into smaller clusters of repetitions with short rest intervals in between, allowing for brief recovery periods during the set.
    +   Example: 3 sets of 10 repetitions performed as 2 clusters of 5 repetitions with 15 seconds rest between clusters and 60 seconds rest between sets.
- Primary research goal: characterize the intensity of TRT vs CRT
- Today's Question: Are there differences in heart rate response between traditional and cluster resistance volume-matched protocols?

## Experimental Design

- Participants: 10 college students
- Cross-over design
    + Each participant completed both the TRT and CRTprotocols... *why?*
    + Randomized (flipped a coin) which protocol was completed first... *why?*
    + One week washout period between protocols... *why?*
    + Measured heart rate, before and after exercise
- This was a class project conducted by Jayva Anderson, Emily Lettenmaier, Sayre Schreiner, and Greta Valaika

## The Data

-  Stored in `TRAD_CLU_group.csv` on Teams in the Jan 7th Activities Module in the Classwork section
-  Columns:
    - ID : Participant identifier
    - Condition: TRAD or CLU (traditional or cluster)
    - Time: PRE or POST (before or after exercise)
    - HR: Heart Rate (in beats per minute)

## Challenge

Make a visualization that addresses the research question.

Best viz wins. Once you're complete post your image in the Activities Channel on Teams.
```{r}
#| echo: false
countdown::countdown(minutes = 30)
```

## For Next Class

-  Complete reflection
-  Complete next two lectures
