---
title: Challenges!
format: revealjs
auto-stretch: false
---

## Today

-   Some challenges to practice your skills!



## Challenge 1A: Parks

- Use the `parks.csv` found in Teams to make this plot.

```{r}
#| echo: false
library(tidyverse)
parks <- read_csv(here::here("Data", "Parks", "parks.csv"))

parks |> 
    filter(city %in% c("New York", "Chicago")) |> 
    ggplot(aes(x = year, y = total_points, fill = city)) +
    geom_col(position="dodge")
```

## Challenge 1B: Parks

- Use the `parks.csv` found in Teams to make this plot.

```{r}
#| echo: false

parks |> 
    filter(city %in% c("New York", "Chicago")) |> 
    select(year, city, total_points) |>
    pivot_wider(
        names_from = city,
        values_from = total_points
    ) |> 
    ggplot(aes(x = `New York`, y = Chicago)) +
    geom_point() + 
    labs(
        x = "New York Total Points",
        y = "Chicago Total Points"
    )
```

## Challenge 1C: Parks

- Use the `parks.csv` found in Teams to make this plot.

```{r}
#| echo: false

parks |> 
    filter(city %in% c("New York", "Chicago", "Los Angeles", "Boise")) |> 
    pivot_longer(
        cols = contains("points"),
        names_to = "Category",
        values_to = "Score"
    ) |> 
    filter(Category != "total_points") |>
    ggplot(aes(x = year, y = Score, fill=Category)) +
    geom_col(position="stack") +
    coord_flip() +
    facet_wrap(~city)
```

## Challenge 2: Valentine's Day

[Generate this plot (without usiung AI):](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-02-13/readme.md)

```{r}
#| echo: false

library(tidyverse)
historical_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/historical_spending.csv')

historical_spending |> 
    select(-PercentCelebrating) |> 
    pivot_longer(
        cols = !Year,
        names_to = "Category",
        values_to = "Spending"
    ) |> 
    ggplot(aes(x = Year, y = Spending, color = Category)) +
        geom_line()
```

## Challenge 3: Pokemon

[Generate the following data frame from the pokemon data (no AI):](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-04-01/readme.md)
```{r}
#| echo: false

library(ggforce)
pokemon_df <- pokemon::pokemon

pokemon_df |> 
    pivot_longer(
        cols = type_1:type_2,
        names_to = "Type_num",
        values_to = "Type"
    ) |> 
    drop_na(Type) |> 
    pivot_longer(
        cols = hp:speed,
        names_to = "Stat",
        values_to = "Value"
    ) |>
    ggplot(aes(y = Type, x = Value, color = generation_id)) +
    geom_sina(size = 0.75, width=0.5) +
    facet_wrap(~Stat) + 
    scale_x_discrete()
```

## Groups for Today

-   Self sort:
    -   Easy: Table 1 (near the windows)
        + Challenge 1A, 1B, 1C
    -   Medium: Table 2 (middle of the room)
        + Challenge 2 and 1A
    -   Hard: Table 3 (near the TVs)
        + Challenge 3
-   Start by sketching a plan on the white board...



## For Next Class

-  Complete reflection
-  Complete next three lectures
