---
title: Uploads and Downloads
subtitle: Lecture 27
---

# File Uploads

## UI: `fileInput`

-   **`fileInput(inputId, label, ...)`**: Creates a file upload control.
-   Let's look at the documentation with `?fileInput`.

```{r}
#| eval: false
fileInput("upload", "Upload a file", multiple = TRUE)
```

## Server: `input$upload` {.smaller}

-   Accessing the input returns a **data frame** with one row per file:
    -   **`name`**: Original filename (e.g., "data.csv").
    -   **`size`**: Size in bytes.
    -   **`type`**: [MIME type](https://en.wikipedia.org/wiki/MIME) (don't usually need to worry about this).
    -   **`datapath`**: Temporary path where the file is saved on the server.
-   **Note**: `datapath` is temporary. Read the data from there immediately.

```{r}
#| eval: false
output$contents <- renderTable({
  req(input$upload)
  read_csv(input$upload$datapath)
})
```

## File Size Limits

-   Default limit is **5 MB**.
-   Increase it by setting an option (usually at the top of `app.R` or `server.R`):

```{r}
#| eval: false
options(shiny.maxRequestSize = 30 * 1024^2) # 30 MB
```

# File Downloads

## UI: `downloadButton`

-   **`downloadButton(outputId, label)`**: A button to initiate a download.
-   **`downloadLink(outputId, label)`**: A link to initiate a download.

```{r}
#| eval: false
downloadButton("downloadData", "Download CSV")
```

## Server: `downloadHandler`

-   **Important**: Use `downloadHandler()` NOT `render*()`.
-   **Arguments**:
    -   **`filename`**: A function that returns the filename (string).
    -   **`content`**: A function that takes a `file` path argument and writes content to it.

```{r}
#| eval: false
output$downloadData <- downloadHandler(
  filename = function() {
    paste("data-", Sys.Date(), ".csv", sep="")
  },
  content = function(file) {
    write.csv(data(), file)
  }
)
```

## Downloading Reports

-   You can also download generated reports (HTML, PDF, Word) using R Markdown or Quarto.
-   This is quite difficult so we won't address it here.

# Wrap Up

## Mechanics

-   **Upload**:
    -   UI: `fileInput()`.
    -   Server: `input$id` (dataframe w/ `datapath`).
-   **Download**:
    -   UI: `downloadButton()`.
    -   Server: `downloadHandler(filename, content)`.

## Do Next

1. Read [Chapter 9: Uploads and Downlaods](https://mastering-shiny.org/action-transfer.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. I reommend working through the Exercises in the textbook.
4. That's it for tonight! See you tomorrow!