---
title: Your First Shiny App
subtitle: Lecture 18
---

# Introduction to Shiny

## What is Shiny?

-   **Shiny** is an R package that makes it easy to build interactive web apps straight from R.
-   Allows you to create standalone apps on a webpage or embed them in Quarto documents.
-   **Key Components**:
    -   **UI (User Interface)**: Defines how your app *looks*.
    -   **Server**: Defines how your app *works*.

## Installation

-   If you haven't installed it yet:

```{r}
#| eval: false
install.packages("shiny")
```

-   Load it in your R session:

```{r}
library(shiny)
```

# Creating an App

## App Structure (`app.R`)

-   The simplest way to create an app is a single file named `app.R`.
-   It contains three main parts:
  1.  **UI object**: The HTML webpage humans interact with.
  2.  **Server function**: The logic that tells the app what to do.
  3.  **`shinyApp` call**: Combines UI and server to run the app.

## Example

```{r}
#| eval: false
library(shiny)

ui <- fluidPage(
  "Hello, world!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

## Running the App

-   **In RStudio/Positron**:
    -   Click the **Run App** button (play icon) in the toolbar.
-   **Stopping**:
    -   Click the **Stop** sign in the console.


## Adding UI Controls {.smaller}

-   Use **Layout Functions** to structure the page (e.g., `fluidPage`).
-   Use **Inputs** to let users provide values.
-   Use **Outputs** to show results.

```{r}
#| eval: false
library(shiny)
ui <- fluidPage(
  selectInput("dataset", label = "Dataset", choices = ls("package:datasets")),
  verbatimTextOutput("summary"),
  tableOutput("table")
)
```

-   `selectInput()`: A dropdown menu.
-   `verbatimTextOutput()`: Placeholder for code/text output.
-   `tableOutput()`: Placeholder for a table.

# Adding Behavior (Server)

## The Server Function

-   Tells Shiny how to build the outputs from the inputs.
-   **Rules**:
    1.  Save objects to `output$ID`.
    2.  Build objects with `render{Type}()`.
    3.  Access inputs with `input$ID`.

## Connecting Inputs to Outputs

```{r}
#| eval: false
server <- function(input, output, session) {
  output$summary <- renderPrint({
    dataset <- get(input$dataset, "package:datasets")
    summary(dataset)
  })

  output$table <- renderTable({
    dataset <- get(input$dataset, "package:datasets")
    dataset
  })
}
```

-   `renderPrint()` pairs with `verbatimTextOutput()`.
-   `renderTable()` pairs with `tableOutput()`.
-   **Reactivity**: Outputs automatically update when `input$dataset` changes.

# Reactivity

## Reducing Duplication

-   In the previous example, we fetched the dataset twice:

```r
dataset <- get(input$dataset, "package:datasets")
```

-   **Don't** use regular variables or functions to solve this (they won't work as expected in the reactive graph).
-   **Do** use **Reactive Expressions**.

## Reactive Expressions {.smaller}

-   Wrap code in `reactive({...})`.
-   Call it like a function: `my_reactive()`.
-   **Benefits**:
    -   Updates only when inputs change.
    -   Caches the result (lazy evaluation) - efficient!

```{r}
#| eval: false
server <- function(input, output, session) {
  # Create a reactive expression
  dataset <- reactive({
    get(input$dataset, "package:datasets")
  })

  output$summary <- renderPrint({
    # Call it like a function
    summary(dataset())
  })

  output$table <- renderTable({
    dataset()
  })
}
```

# Summary

## Recap

-   **UI**: `fluidPage()`, `input*()`, `*Output()`.
-   **Server**: `function(input, output, session)`.
-   **Wiring**:
    -   `output$ID <- render*({ code using input$ID })`
-   **Reactivity**:
    -   `reactive()` calculates values and caches them.
    -   Outputs automatically "react" to input changes.
-   **Next Steps**: Basic UI, Basic Reactivity (Chapters 2 & 3).

## Do Next

1. Read [Chapter 1: Your First Shiny App](https://mastering-shiny.org/basic-app.html) from Mastering Shiny.
2. There's NO recitation Gem for this textbook. I recommend doing the following:
  - Create your own NotebookLM for this text book. Upload the Chapters as we go along and the slides we use (don't forget to use press`e` before imputting the link to the slides).
  - Use the NotebookLM as a tutor!
  - You can also use the NotebookLM as context within your Gemini chats.
3. Move on to Lecture 19!