---
title: "Unsupervised Learning"
subtitle: "Lecture 33"
---

## Unsupervised Learning {.smaller}

-   **Machine Learning**: using computers and algorithms to learn from data.
-   Two types of machine learning problems are:
  -   **Supervised Learning**: you have a **target** variable that you're trying to predict
  -   **Unsupervised Learning**: you don't have a target variables and you're trying to extract patterns from data  
  -   Two common types of unsupervised learning are:
    -   **Dimensionality reduction**: find a lower-dimensional representation of the data that captures most of the variation (PCA)
    -   **Clustering**: find subgroups of observations that are similar to each other

## K-means Clustering

K-means clustering is a simple and popular clustering method. It works by partitioning the data into K distinct, non-overlapping clusters.

The `tidyclust` package provides a `tidymodels`-like interface for clustering.

```{r}
#| message: false
library(tidymodels)
library(tidyclust)
library(tidyverse)
```

We will use the `USArrests` dataset. It contains statistics, in arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states in 1973. It also gives the percent of the population living in urban areas.

```{r}
USArrests |>  head()
```

## K-means with tidyclust

Let's do a K-means clustering with 2 clusters. We create a `k_means()` model specification and set the engine.

```{r}
kmeans_spec <- k_means(num_clusters = 2) |>
  set_engine("stats")

kmeans_spec
```

Then we can fit the specification. We are using a formula `~ .` to specify that we want to use all variables.

```{r}
kmeans_fit <- kmeans_spec |> 
  fit(~ ., data = USArrests)

kmeans_fit
```

## Understanding the result

We can use `tidy()` to get a tibble with information about each of the clusters. This includes the size of each cluster, the location of the cluster centroid, and the within-cluster sum of squares.

```{r}
tidy(kmeans_fit)
```

We can also use `augment()` to get the cluster assignment for each observation.

```{r}
augment(kmeans_fit, USArrests) |>
  slice(1:5)
```

## Visualizing Clusters

With the cluster assignments, we can now visualize the clusters.

```{r}
#| code-fold: true
kmeans_fit |>
  augment(USArrests) |>
  ggplot(aes(x = Murder, y = Assault, color = .pred_cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering (K=2)") +
  theme_minimal()
```

## Visualizing Clusters PCA

With the cluster assignments, we can now visualize the clusters.

```{r}
#| code-fold: true
pca_rec <- recipe(~ ., data = USArrests) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 2) |> 
  prep()

US_Arrests_PCA <- bake(pca_rec, new_data = USArrests)

kmeans_fit |>
  augment(USArrests) |>
  bind_cols(US_Arrests_PCA) |> 
  ggplot(aes(x = PC1, y = PC2, color = .pred_cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering (K=2)") +
  theme_minimal()
```

## Tuning the number of clusters

A common question in K-means clustering is how to choose the number of clusters K. A common way to do this is to try different values of K and see which one gives the best result.

We can use `tune_cluster()` to do this. We first create a `k_means()` specification with a `num_clusters` argument that we want to tune.


# Hierarchical Clustering

Hierarchical clustering is another popular clustering method. It creates a hierarchy of clusters that can be visualized as a dendrogram.

Let's do a hierarchical clustering with `tidyclust`.

```{r}
hc_spec <- hier_clust() |>
  set_engine("stats")

hc_fit <- fit(hc_spec, ~ ., data = USArrests)

hc_fit
```

## Dendrogram

The result of hierarchical clustering is often visualized as a dendrogram. We can use the `factoextra` package to do this.

```{r}
#| message: false
library(factoextra)

fviz_dend(hc_fit$fit)
```

## Linkage

In hierarchical clustering, we need to specify a linkage method, which determines how the distance between clusters is calculated. Common linkage methods are:

-   `complete`: The distance between two clusters is the maximum distance between any two points in the two clusters.
-   `average`: The distance between two clusters is the average distance between all pairs of points in the two clusters.
-   `single`: The distance between two clusters is the minimum distance between any two points in the two clusters.

We can specify the linkage method in `hier_clust()`.

```{r}
#| eval: false
hier_clust(linkage_method = "average")
```

# Summary

- **Clustering**: Finding groups in data.
- **K-Means**: Partitions data into K clusters. Need to choose K.
- **Hierarchical Clustering**: Creates a hierarchy of clusters.
- **`tidyclust`**: A `tidymodels`-like interface for clustering.
- `tune_cluster()` can be used to tune the number of clusters.