---
title: Clustering with Tidyclust
subtitle: Lecture 33
---

# Unsupervised Learning

## Finding Groups

- **Supervised Learning**: We have a target variable (Y) to predict (Regression, Classification).
- **Unsupervised Learning**: No target. We want to find *structure* in the data.
- **Clustering**: Grouping similar observations together.
- **Techniques**:
  -   K-Means Clustering
  -   Hierarchical Clustering

## The `tidyclust` Package

- Builds on `tidymodels`.
- Provides a consistent interface for clustering models.
- **Workflow**:
  1.  Specification: `k_means()` or `hier_clust()`.
  2.  Engine: `set_engine()`.
  3.  Fit: `fit()`.

```{r}
library(tidymodels)
library(tidyclust)
library(tidyverse)
library(palmerpenguins)

# Prepare Data (Numeric, Scaled, No NA)
penguins_clean <- penguins |>
  drop_na() |>
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  scale() |>
  as_tibble()
```

# K-Means Clustering

## Intuition

- **Goal**: Partition data into $K$ clusters.
- **Algorithm**:
  1.  Initialize $K$ centroids randomly.
  2.  Assign each point to the closest centroid.
  3.  Update centroids (mean of assigned points).
  4.  Repeat until convergence.

## Fitting K-Means

```{r}
# 1. Specification (Assume K=3)
kmeans_spec <- k_means(num_clusters = 3) |>
  set_engine("stats")

# 2. Fit
kmeans_fit <- kmeans_spec |>
    fit(~ ., data = penguins_clean)


kmeans_fit
```

## Visualizing Clusters

- We can explore the cluster assignments.

```{r}
# Extract Assignments
# extract_cluster_assignment(kmeans_fit) # This extracts just the vector

# Visualize
kmeans_fit |>
  extract_cluster_assignment(penguins_clean) |>
  bind_cols(penguins_clean) |>
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = .cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering (K=3)") +
  theme_minimal()
```

## Choosing K (Elbow Method)

- How do we know K=3 is best?
- **Total Within-Cluster Sum of Squares (WSS)**: Measure of compactness (lower is better).
- Look for the "Elbow" in the WSS vs K plot.

```{r}
#| eval: false
# Note: tune_cluster() supports tuning K, but for simplicity:
# Iterate manually or use standard loop for elbow plot
# (tidyclust makes simple fitting easy, tuning is advanced workflow)
```

# Hierarchical Clustering

## Intuition

- Builds a hierarchy of clusters.
- **Dendrogram**: Tree diagram showing relationships.
- No need to pre-specify K!
- **Linkage**: How to define distance between clusters (Complete, Single, Average).

## Fitting Hierarchical Clustering

```{r}
hc_spec <- hier_clust(linkage_method = "complete") |>
  set_engine("stats")

hc_fit <- hc_spec |>
  fit(~ ., data = penguins_clean)

hc_fit
```

## Visualizing Dendrograms

- `tidyclust` (and `dendextend` underneath) makes plotting easy.

```{r}
# In tidyclust 0.2.0+, plotting might need specific helper or extraction
# Standard plot method often works on the engine object
# plot(hc_fit$fit)

# Using as.dendrogram if needed
# hc_fit$fit |> as.dendrogram() |> plot()
```

*(Note: Simple base R plot of the underlying `hclust` object is often easiest for slides)*

```{r}
plot(hc_fit$fit, main = "Dendrogram of Penguins", labels = FALSE, xlab = "")
rect.hclust(hc_fit$fit, k = 3, border = "red")
```

## Cutting the Tree

- We can "cut" the tree to get $K$ clusters.

```{r}
# Re-fit with num_clusters if we want prediction-like behavior or use cutree on engine
# tidyclust handles this via prediction usually.
```

# Summary

## Summary

- **Clustering**: Finding groups without labels.
- **K-Means**: Partition into K groups. Good for spherical clusters.
- **Hierarchical**: Tree structure. Good for visualizing relationships.
- **Tidyclust**: Brings tidy syntax to clustering (`k_means()`, `hier_clust()`, `fit()`).
- **Visualization**:
  -   Scatterplots colored by Cluster.
  -   Dendrograms.
