---
title: "Unsupervised Learning"
subtitle: "Lecture 33"
---

## Unsupervised Learning {.smaller}

-   **Machine Learning**: using computers and algorithms to learn from data.
-   Two types of machine learning problems are:
    -   **Supervised Learning**: you have a **target** variable that you're trying to predict
    -   **Unsupervised Learning**: you don't have a target variables and you're trying to extract patterns from data
-   Two common types of unsupervised learning are:
    -   **Dimensionality reduction**: find a lower-dimensional representation of the data that captures most of the variation (PCA)
    -   **Clustering**: find subgroups of observations that are similar to each other

## K-means Clustering

-   A simple and popular clustering method.
-   **Goal**: Partition data into $K$ distinct, non-overlapping clusters.

## How it Works

-   [An iterative algorithm](https://019bffe2-edb8-4bc6-f8aa-0d38338bc8a1.share.connect.posit.cloud/) :
    1.  **Initialize**: Randomly select $K$ data points as initial cluster centroids.
    2.  **Assign**: Assign each data point to the *nearest* centroid, forming K clusters.
    3.  **Update**: Recalculate the centroid of each cluster (the mean of all points in the cluster).
    4.  **Repeat**: Repeat "Assign" and "Update" steps until cluster assignments stop changing.

## `tidyclust`

The `tidyclust` package provides a `tidymodels`-like interface for clustering.

```{r}
#| message: false
library(tidymodels)
library(tidyclust)
library(tidyverse)
```

## The Data: `USArrests`

-   Contains statistics for the 50 US states in 1973:
    -   Arrests per 100,000 residents for Assault, Murder, and Rape.
    -   Percent of the population living in urban areas.

```{r}
USArrests |> head()
```

## K-means with `tidyclust`

-   Create a `k_means()` model specification.
-   Set the number of clusters (`k=2` for this example).
-   Set the engine to `"stats"`.

```{r}
kmeans_spec <- k_means(num_clusters = 2) |>
  set_engine("stats")

kmeans_spec
```

## K-means with `tidyclust`

-   Then we can fit the specification.
-   We are using a formula `~ .` to specify that we want to use all variables.

```{r}
kmeans_fit <- kmeans_spec |>
  fit(~., data = USArrests)

kmeans_fit
```

## Understanding the result {.smaller}

-   `tidy()` gives information about each cluster:
    -   size of each cluster
    -   cluster centroid locations
    -   within-cluster sum of squares
-   `augment()` gives the cluster assignment for each observation.

```{r}
tidy(kmeans_fit)
```

```{r}
augment(kmeans_fit, USArrests) |>
  head()
```

## Visualizing Clusters

-   With the cluster assignments, we can now visualize the clusters.

```{r}
#| code-fold: true
kmeans_fit |>
  augment(USArrests) |>
  ggplot(aes(x = Murder, y = Assault, color = .pred_cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering (K=2)") +
  theme_minimal()
```

## Visualizing Clusters PCA

-   We can also visualize the clusters in the PCA space.

```{r}
#| code-fold: true
pca_rec <- recipe(~., data = USArrests) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 2) |>
  prep()

US_Arrests_PCA <- bake(pca_rec, new_data = USArrests)

kmeans_fit |>
  augment(USArrests) |>
  bind_cols(US_Arrests_PCA) |>
  ggplot(aes(x = PC1, y = PC2, color = .pred_cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering (K=2)") +
  theme_minimal()
```

## Tuning the number of clusters

-   How to choose the number of clusters K?
-   A common way is to try different values of K and see which one gives the best result.
-   We can use `tune_cluster()` to do this. Not this class.

# Hierarchical Clustering

## Hierarchical Clustering

-   Another popular clustering method.
-   Creates a hierarchy of clusters, often visualized as a dendrogram.

## How it Works (Agglomerative)

-   [It's a "bottom-up" approach](https://019bffe2-edb8-4bc6-f8aa-0d38338bc8a1.share.connect.posit.cloud/):
    1.  **Start**: Each data point begins in its own cluster.
    2.  **Merge**: Find the two "closest" clusters and merge them into a single cluster.
    3.  **Repeat**: Repeat the "Merge" step until all data points are in a single cluster.
-   The history of merges forms the hierarchy.

## Using `tidyclust`

Let's do a hierarchical clustering with `tidyclust`.

```{r}
hc_spec <- hier_clust() |>
  set_engine("stats")

hc_fit <- fit(hc_spec, ~., data = USArrests)

hc_fit
```

## Dendrogram

-   The result of hierarchical clustering is often visualized as a dendrogram.
-   We can use the `factoextra` package to do this.

```{r}
#| message: false
library(factoextra)

fviz_dend(hc_fit$fit)
```

## Linkage {.smaller}

-   In hierarchical clustering, we need to specify a linkage method, which determines how the distance between clusters is calculated.
-   Common linkage methods are:
    -   `complete`: The distance between two clusters is the maximum distance between any two points in the two clusters.
    -   `average`: The distance between two clusters is the average distance between all pairs of points in the two clusters.
    -   `single`: The distance between two clusters is the minimum distance between any two points in the two clusters.

We can specify the linkage method in `hier_clust()`.

```{r}
#| eval: false
hier_clust(linkage_method = "average")
```

# Wrap-Up

## Summary

-   **Clustering**: Finding groups in data.
-   **K-Means**: Partitions data into K clusters. Need to choose K.
-   **Hierarchical Clustering**: Creates a hierarchy of clusters.
-   **`tidyclust`**: A `tidymodels`-like interface for clustering.
-   `tune_cluster()` can be used to tune the number of clusters.

## Do Next

1. (Optional)Read [Chapter 12: Unsupervised Learning](https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/12-unsupervised-learning.html) from [ISLR](https://www.statlearning.com/).
2. There's NO recitation Gem for this textbook but I recommend creating your own and adding the textbook chapter and these slides.
3. Move on to Lecture 34!