---
title: "Logical Vectors"
subtitle: "Lecture 9"
---

```{r}
#| label: setup
#| message: false
#| echo: false

# load packages
library(tidyverse)
library(nycflights13)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)
```

# Introduction

## Logical Vectors

- A logical vector is a vector that contains only `TRUE`, `FALSE`, and `NA` (missing) values.
- They are rarely in your raw data, but you will create and use them in almost every analysis.
- Logical vectors are the foundation of filtering and conditional transformations.
- In CS Logical Vectors are often called "Boolean" vectors.

# Comparisons

## Creating Logicals

- You can create logical vectors with comparison operators:
  - `>` (greater than), `<` (less than)
  - `>=` (greater than or equal to), `<=` (less than or equal to)
  - `==` (equal to), `!=` (not equal to)

```{r}
#| echo: true
x <- c(1, 2, 3, 5, 8, 13)
x > 4
```

## Comparisons with `filter()`

- Most often, you'll use comparisons inside `filter()` to select rows.
- Let's use the `flights` dataset to find flights that were delayed on arrival.

## Floating Point Comparison

- Be careful when using `==` with numbers. Computers can have small rounding errors.
- Use `near()` from the `dply` package to compare two numbers for "close enough" equality.
- Let's see an example.

# Missing Values

## `NA`

- `NA` represents a missing value.
- `NA` is "contagious": almost any operation involving `NA` will produce `NA`.
- You can't use `==` to find missing values.

```{r}
#| echo: true
NA > 5
NA == 10
NA == NA
```

## `is.na()`

- Use the `is.na()` function to test if a value is missing.

```{r}
#| echo: true
x <- c(1, NA, 3)
is.na(x)
```

## Filtering and `NA`

- `filter()` only includes rows where the condition is `TRUE`. It excludes `FALSE` and `NA` values.
- If you want to include missing values in your filter, you must ask for them explicitly.

## Removing Missing Values

- Use `drop_na` to remove rows with missing values.
- Specify the columns you want to remove from.
- Avoid dropping more rows than necessary.

## Practice!

Let's do some practice!

# Boolean Algebra

## Combining Conditions

- You can combine multiple logical conditions using boolean operators:
  - `&` (and)
  - `|` (or)
  - `!` (not)

## Examples

-  Let's select flights that were delayed on both departure and arrival.
-  Let's select flights that were delayed on either departure or arrival.
-  Let's select flights that were not delayed on either departure or arrival.
-  Let's select flights that were from November or December.

## `%in%`

- A useful shortcut for `x == a | x == b | x == c` is `x %in% c(a, b, c)`.
- It checks if each value in the left vector is present in the right vector.

# Summarizing Logicals

## Numeric Summaries

- When you use a logical vector in a numeric context, `TRUE` becomes `1` and `FALSE` becomes `0`.
- `sum()`: counts the number of `TRUE`s.
- `mean()`: calculates the proportion of `TRUE`s.

## Practice!

-  Did Delta have any departure delays on Christmas Day? If so how many?
-  How many departure delays did AA, DL, and UA have on Christmas Day?
-  Plot the total number of departure delays per month. Per day.

# Conditional Transformations

## `if_else()`

- `if_else(condition, value_if_true, value_if_false)`
- All output values must be of the same type.

```{r}
#| echo: true
x <- c(-3, -1, 0, 1, 3)
if_else(x > 0, "positive", "negative or zero")
```

## `case_when()`

- For multiple conditions, `case_when()` is easier to read than nested `if_else()`.
- `case_when(condition1 ~ value1, condition2 ~ value2, ...)`
- The first `TRUE` condition wins.

```{r}
#| echo: true
#| code-fold: true
case_when(
  x < 0 ~ "negative",
  x == 0 ~ "zero",
  x > 0 ~ "positive"
)
```

## Example: `case_when()`

- Let's create a human-readable "season" variable.

# Wrap Up

## Logical Vectors

- Create with comparisons: `<`, `==`, `%in%`, `is.na()`.
- Combine with boolean operators: `&`, `|`, `!`.
- Summarize with `any()`, `all()`, `sum()`, `mean()`.
- Use for conditional transformations with `if_else()` and `case_when()`.
- Logical vectors are a fundamental tool for data manipulation in R.

## Do Next

1. Read [Chapter 12: Logicals](https://r4ds.hadley.nz/logicals.html) from r4ds.
2. Open the Recitation Gem and say "Provide me practice problems for Chapter 12" or work through some of the exercises in the text.
3. Move on the Lecture 11!
